Molecular dynamics simulation of synchronization in driven particles

### 2020-11-19 manuscript/manuscript_ajp_2020.tex
updated text

  %What problem did you study and why is it important?
  We study synchronization of interacting particles
  confined to a narrow channel 
  driven by an external applied force.  
  %What methods did you use?
  Using 
  numerical simulations
  we control the particle interactions,
  external force and environment
  in order to mimick experimental studies of driven colloidal particles
  confined by light-fields.
  The molecular dynamics simulations
  model the particle dynamics 
  using overdamped equations of motion suitable for a viscous suspension
  of microscopic particles.
  %What were your principal results?
  We observe
  particle synchronization under a variety of conditions,
  including static and dynamic patterns formed on
  periodic and quasiperiod substrates
  and the propagation of 
  high density kinks.
  %What conclusions have you drawn from these results?
  We demonstrate the transition from
  trapped to sliding dynamics in
  quasiperiodic landscapes differs
  from that of a periodic landscape by [TBD],
  and that kinks of high density propagate
  [more/less ?  ] when [what?].
  We also explore 
  the limits of applied force that
  cause the dynamics to transition to chaotic behavior.
  We include sample code
  and exercises for students
  that include 
  opportunties
  to reproduce our results and propose
  new numerical experiments.
  With only a few particles in two-dimensions,
  the simulation runs quickly,
  making this an appropriate model for undergraduates to explore.

@@

\section{Introduction} %
%
%Outline
%------------------------------------------------------------------
%synchronization = widest audience possible
Synchronization is a universal phenomena
that can be modeled with numerical studies of confined,
driven particles.
%Christiaan Huygens 
%[has been adapted into many useful technologies...]
Examples of synchronization occur
in many naturally oscillating systems,
from Huygens 1665 observation of
in-phase oscillation of pendulum clocks \cite{Pikovsky2003},
to fireflies signaling in time with a with a pulsed LED \cite{},
to simultaneous flapping of birds wings \cite{},
to neuronal firing in cardiac muscle \cite{} and brain tissue \cite{}.

%phase locking
Dynamical mode locking is a controlled 
synchronization phenomena in 
which an external driving force
causes
%the natural oscillation frequencies
%in
weakly coupled oscillators to come into phase.
Mode locking is often observed in quantum electronic
devices such as Josephson junctions \cite{}
and phase-locked current loops \cite{}.
Instead of ohmic current-voltage (IV) relationships,
such devices exhibit stepped regions
such as Shapiro steps \cite{} [explain].

Recently studies of colloidal particles confined
in optical traps have
been used to examine the microscopic dynamics
of mode-locking when subject to external driving force \cite{juniper2015}.
[this is important because...]
%experimental - colloids specifically
Colloidal particles trapped in light fields have proven
a particularly useful medium for studying complex dynamical behaviors.  %cite 
The relatively large size of the colloids
and ease in control has lead to a rich array of 
experimental results. %THIS IS A TERRIBLE SENTENCE.
Such studies are considered model systems
for experimental systems relatively hard to access and visualize,
such as cold atoms or electron gases. %ELABORATE AND CITE

@@

%external forces
It is often useful to study the
dynamics of particles subject to 
an applied external force,
for instance the flow of charges in a conductor,
or cells responding to a chemical gradient.
The external force 
increases the diversity of dynamical behaviors,
and can cause particles to flow in
a variety of non-linear complex patterns
that may be 
synchronized.
Disordered chaotic dynamics are also possible,
where irregular, unpredictable time evolution of
nonlinear systems and occurs in mechanical oscillators \cite{chaos}.

@@

@@

%describe types of physical systems - particle types
In this work,
we perform 
numerical simulations of confined, driven particles
to model a variety of physical phenomena.
%----------------------------------------------------------------------------
%paper outline
In the following paper we describe
our molecular dynamics model in Section~\ref{sec:MD}.
%including various  and confining environments.
We include code to simulate
and visualize the dynamics in this section
and supplementary material.
In Section~\ref{sec:results} we summarize
our results,
including synchronized motion of a single confined particle
driven across a periodic landscape in 
Section~\ref{sec:one} and
multiple interacting particles
in Section~\ref{sec:sync},
including stationary propagation of high density kinks
in Sec.~\ref{sec:kink}.
%we
%demonstrate how uniform environments and applied forces
%and create synchronized flow patterns.
We present these results using standard tools of non-linear oscillators
such phase diagrams of particle velocity versus position.
In Section~\ref{sec:quasiperiod},
we show how an aperiodic landscape modifies the particle dynamics.
Finally we explore the transition to chaotic dynamics in 
Section~\ref{sec:chaos}
and conclude 
in Section~\ref{sec:conclusion}. 
In each section
we suggest exercises for interested students,
as summarized in Section~\ref{sec:problems}.

@@

\begin{table}[h!]
\centering
\caption{Simulation units defined in the code.}
\begin{ruledtabular}
\begin{tabular}{c c p{5cm}}
Quantity & Unit \\
\hline
length &  $a_0$ \\
viscosity/damping constant & $\eta$ \\
dimensionless interaction strength & $q$ \\
effective colloidal charge & $Z^*$ \\
solvent dielectic constant & $\epsilon$ \\
time &  $\eta / E_0$ \\
velocity &  $\eta E_0 / a_0$ \\
force & $E_0 / a_0$\\
electric potential & $V(r_{ij}) = q^2 E_0/r_{ij}$ \\
energy & $E_0 = {Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

@@

 We model several enviroments
 %creating a landscape of potential minima and maxima
 %that modify the force on a particle as a function of position
 $\vec{F}_{l}(\vec{r}) = -\nabla V_l(\vec{r}).$
 The landscape potential $V(\vec{r}) $ are static 
 with fixed minima and maxima
 that are periodic or quasi-periodic,
 as described in Sec.~\ref{sec:results}. 
 One example of a periodic landscape is
 %that used in Sec.~\ref{sec:one}
 where 
 \begin{equation}
   V_l(y) = V_0 \cos{(N_p \pi y / L)}
 \end{equation}
 where $N_p$ are the number of periods,
 and $V_0$ is an adjustable parameter
 to set the depth of the troughs.

@@

 We model particle interactions with
 the Yukawa potential $\vec{F}_{ij} = -\nabla V_{ij}(r_{ij})$,
 \begin{equation}
   V_{ij}(r_{ij}) = \frac{E_0}{r_{ij}} e^{-\kappa r_{ij}},
 \end{equation}
 where $r_{ij} = |\vec{r}_i - \vec{r}_j|$.
 This is a
 screened Coulomb potential
 $E_0=2$ scales strength of repulsion
 where $E_0 = k q_1 q_2$  [CHECK scaling/units].
 $\kappa = 1/R_0$ is the screening parameter 
that accounts for the lengthscale at
which many particle
interactions and local environment
that reduces the interaction range of individual particles.
We fix $R_0$ to be unity in our simulation units.

@@

The hopping pattern of the driven particle is periodic,
and could be achieved with a variety of
driving forces.  %, as we will demonstrate in Fig.~\ref{fig:1}
In Fig.~\ref{fig:1}
we show the result increasing the 
force to the particle as a function of time.  
%increased DC
We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$
at a rate of $\Delta F_{DC} = 0.001$ every $\Delta t = 4000$ integer timesteps
or $40$ time units.
%this seems fast, but it is slow enough to estimate
%the velocity accurately.
%this is not be so for a multiple particle system!
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.

>>>>

  %What problem did you study and why is it important?
  We study synchronization of interacting particles
  confined to a narrow channel 
  driven by an external applied force.  
  %What methods did you use?
  Using 
  numerical simulations
  we control the particle interactions,
  external force and confining environment
  in order to mimick experimental studies of driven colloidal particles
  confined by light-fields.
  The molecular dynamics simulations
  %model the particle dynamics 
  use overdamped equations of motion suitable for a viscous suspension
  of microscopic particles
  to model the dynamics.
  %What were your principal results?
  We observe
  particle synchronization under a variety of conditions,
  including static and dynamic patterns formed on
  periodic and quasiperiod substrates
  and the propagation of 
  high density regions of particles,
  or kinks.
  %What conclusions have you drawn from these results?
  We demonstrate the transition from
  trapped to sliding dynamics in
  quasiperiodic landscapes differs
  from that of a periodic landscape by [TBD],
  and that kinks of high density propagate
  [more/less ?  ] when [what?].
  We also explore 
  the limits of applied force that
  cause the dynamics to transition to chaotic behavior.
  We include sample code
  and exercises for students
  %that include
  with 
  opportunties
  to reproduce our results and propose
  new numerical experiments.
  With only a few particles in two-dimensions,
  the simulation runs quickly,
  making this an appropriate model for undergraduates to explore.

@@

\section{Introduction} %
%
%Outline
%------------------------------------------------------------------
%synchronization = widest audience possible
Synchronization is a universal phenomena
that can be modeled with numerical studies of confined,
driven particles.
%Christiaan Huygens 
%[has been adapted into many useful technologies...]
Examples of synchronization occur
in many naturally oscillating systems,
from Huygens 1665 observation of
in-phase oscillation of pendulum clocks \cite{Pikovsky2003},
to fireflies signaling in time with a pulsed light emitting diode \cite{},
to coordinated flapping of birds wings \cite{},
to simultaneous neuronal firing in cardiac muscle \cite{}
and brain tissue \cite{}.

%phase locking
Dynamical mode locking is a controlled 
synchronization phenomena in 
which an external driving force
causes
%the natural oscillation frequencies
%in
weakly coupled oscillators to come into phase.
Mode locking is often observed in quantum electronic
devices such as Josephson junctions \cite{}
and phase-locked current loops \cite{}.
Instead of ohmic current-voltage (I-V) relationships,
such devices exhibit stepped regions
such as Shapiro steps \cite{} [explain].

Recent experiments of colloidal particles confined
in optical traps
subject to external driving forces 
have
been used to examine the microscopic dynamics
of mode-locking \cite{juniper2015}.
%experimental - colloids specifically
Colloidal particles trapped in light fields are %have proven
a particularly useful medium for studying complex dynamical behaviors.  %cite 
The relatively large size of the colloids (micrometers)
makes colloids easy to control and image,
and the interaction forces between colloids can be modified
by tuning the chemistry of the suspending fluid or surface ligands
of the colloidal particles \cite{Grier2003}.
This ease in control of colloids has lead to a rich array of 
experimental results  
considered models
for experimental systems relatively hard to access and visualize,
such as cold atoms or electron gases. %ELABORATE AND CITE

@@

%external forces
The
dynamics of particles subject to 
an applied external force apply to many physical systems.
For instance the flow of charges in a conductor,
or cells responding to a chemical gradient.
The external force 
increases the diversity of dynamical behaviors,
and can cause particles to flow in
a variety of non-linear complex patterns
that may be 
synchronized.
Disordered chaotic dynamics are also possible,
where irregular, unpredictable time evolution of
nonlinear systems and occurs in mechanical oscillators \cite{chaos}.

@@

Numerical modeling of colloids can provide mechanistic insight
that can be difficult to achieve in experimental conditions
where Brownian motion and other sources of noise dominate.
[elaborate]

@@

%describe types of physical systems - particle types
In this work,
we perform 
numerical simulations of confined, driven particles
to model a variety of physical phenomena.
%----------------------------------------------------------------------------
%paper outline
In the following paper we describe
our molecular dynamics model in Section~\ref{sec:MD}.
%including various  and confining environments.
We include code to simulate
and visualize the dynamics in this section
and supplementary material.
In Section~\ref{sec:results} we summarize
our results,
including synchronized motion of a single confined particle
driven across a periodic landscape in 
Section~\ref{sec:one} and
multiple interacting particles
in Section~\ref{sec:sync},
including stationary propagation of high density kinks
in Sec.~\ref{sec:kink}.
%we
%demonstrate how uniform environments and applied forces
%and create synchronized flow patterns.
We present these results using standard tools of non-linear oscillators
such phase diagrams of velocity versus position.
In Section~\ref{sec:quasiperiod},
we show how an aperiodic landscape modifies the particle dynamics.
Finally we explore the transition to chaotic dynamics in 
Section~\ref{sec:chaos}
and conclude 
in Section~\ref{sec:conclusion}. 
In each section
we suggest exercises for interested students,
as summarized in Section~\ref{sec:problems}.

@@

We model particle interaction forces
$\vec{F}_{ij} = -\nabla V_{ij}(r_{ij})$ 
with
the Yukawa potential energy 
\begin{equation}
  V_{ij}(r_{ij}) = \frac{E_0}{r_{ij}} e^{-\kappa r_{ij}},
\end{equation}
where particle $i$ and $j$ are distance
$r_{ij} = |\vec{r}_i - \vec{r}_j|$ apart.
This 
screened Coulomb potential
%$E_0=2$ scales strength of repulsion
is scaled in terms of energy unit $E_0$
defined in Table ~\ref{tab:1}.
%$ = k q_1 q_2$. %[CHECK scaling/units].
$\kappa = 1/R_0$ is the screening parameter 
that describe the lengthscale at
which particles interact.
We fix $R_0$ to be $a_0$.
In experiments charge screening is observed
due to ions in the suspending fluid and
the charges of surrounding particles
which
reduces the interaction range of individual particles.
This allows the numerical models to run more efficiently
since particles do not need to interact over long range.
[explain particle in cell models].

\begin{table}[h!]
\centering
\caption{Simulation units defined in the code.}
\begin{ruledtabular}
\begin{tabular}{c c p{5cm}}
Quantity & Unit \\
\hline
length &  $a_0$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
dimensionless interaction strength & $q$ \\
effective colloidal charge & $Z^*$ \\
solvent dielectic constant & $\epsilon \epsilon_0$\\
force & $E_0 / a_0$\\
viscosity/damping constant & $\eta$ \\
time &  $\eta / E_0$ \\
velocity &  $ E_0 / \eta a_0$ \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

@@

We use several model enviroments to confine the particles,
again assuming the confining force arises from a potential function 
 %creating a landscape of potential minima and maxima
 %that modify the force on a particle as a function of position
 $\vec{F}_{l}(\vec{r}) = -\nabla V_l(\vec{r}).$
 The landscape potential $V(\vec{r}) $ are static 
 with fixed minima and maxima
 that are periodic or quasi-periodic,
 as described in Sec.~\ref{sec:results}.

 One example of a periodic landscape is
 %that used in Sec.~\ref{sec:one}
 where 
 \begin{equation}
   V_l(y) = V_0 \cos{(N_p \pi y / L)}
 \end{equation}
 where $N_p$ are the number of periods,
 and $V_0$ is an adjustable parameter
 to set the depth of the troughs.

@@

@@

The hopping pattern of the driven particle is periodic,
and could be achieved over a range of
driving forces.  %, as we will demonstrate in Fig.~\ref{fig:1}
We explore the ranges of periodic hopping patterns
by increasing $F_{DC}$ as a function of time,
as shown in 
%
Fig.~\ref{fig:1}.
%we show the result increasing the 
%force to the particle as a function of time.  
%increased DC
We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$
at a rate of $\Delta F_{DC} = 0.001$ every $\Delta t = 4000$ integer timesteps
or $40$ time units.
%this seems fast, but it is slow enough to estimate
%the velocity accurately.
%this is not be so for a multiple particle system!
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.
